Elephant Seal Age JAE Intraspecific Variation

### 2024-05-23 paper/paperv3.qmd
clean up reproductive senescence model

First, address detection probability. We do resights near-daily during breeding and molting seasons, so while we may miss an individual on a given day, it is unlikely that we miss a marked individual throughout the entire breeding or molting season.

```{r}
#| label: detection-prob

full <- semi_join(resight, sealdat, "animalID") %>% 
  mutate(fiscalyear = ifelse(yday > 200, calyear + 1, calyear)) %>% 
  group_by(animalID, fiscalyear) %>% 
  count(fiscalyear) %>% 
  mutate(year = as.numeric(fiscalyear)) %>% 
  filter(year > 1987)

full %>% 
  group_by(year) %>% 
  summarize(across(n, list(median = median, 
                           q25 = \(x) quantile(x, 0.25), 
                           q75 = \(x) quantile(x, 0.75)))) %>% 
  ggplot(aes(year, n_median)) +
  geom_ribbon(aes(ymin = n_q25, ymax = n_q75), fill = "grey80") +
  geom_line() +
  geom_hline(yintercept = 3, linetype = "dashed", color = "firebrick") +
  labs(y = "Resights per year") + 
  theme_classic()

#some animals have multiple rows because they have multiple pups associated with them...
sealdat %>% 
  group_by(animalID, year) %>% 
  mutate(n = n()) %>% 
  filter(n > 2)

skips <- sealdat %>% 
  distinct(animalID, year) %>% 
  group_by(animalID) %>% 
  summarize(n_obs = n(),
            obs_window = max(year) - min(year) + 1,
            skips = obs_window - n_obs) %>% 
  group_by(obs_window, skips) %>% 
  summarize(ind = n(), .groups = "drop")

resights_per_year <- full %>% 
  group_by(year) %>% 
  summarize(median_resights = median(n))

skips_per_year <- skips

skip_frac <- sealdat %>% 
  distinct(animalID, year) %>% 
  group_by(animalID) %>% 
  summarize(firstobs = min(year),
            lastobs = max(year)) %>% 
  mutate(n = lastobs - firstobs + 1) %>% 
  uncount(n) %>% 
  group_by(animalID) %>% 
  mutate(year = firstobs + seq(n()) - 1) %>% 
  ungroup() %>% 
  left_join(transmute(sealdat, animalID, year, skip = 0), 
            by = c("animalID", "year")) %>% 
  mutate(skip = ifelse(is.na(skip), 1, 0)) %>% 
  group_by(year) %>% 
  summarize(skip_frac = sum(skip) / n())

resights_per_year %>% 
  left_join(skip_frac, by = "year") %>% 
  ggplot(aes(median_resights, skip_frac)) +
  geom_point(aes(fill = year), shape = 21, color = "grey20", size = 2) +
  ggrepel::geom_label_repel(aes(label = year), max.overlaps = 8) +
  scale_y_continuous("Skip fraction", labels = scales::percent) +
  scale_fill_distiller(palette = "PuRd") +
  labs(x = "Median resights") +
  theme_classic()

#mark recapture model is necessary? could have bred at another colony. Many skips!!

ggplot(skips, aes(obs_window, skips, size = ind)) +
  geom_point()

#

@@

We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).

```{r}
#| label: age_breeding_mod

breeding_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  data = sealdat, 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

breed_pred <- ggpredict(breeding_mod, terms = "age10 [all]") %>% 
  as_tibble() %>% 
  mutate(age = x * 10)

ranef_year <- ranef(breeding_mod)$year_fct %>% 
  as_tibble(rownames = "year_fct") %>% 
  arrange(`(Intercept)`) %>% 
  slice(1, floor(nrow(.) * c(0.25, 0.5, 0.75, 1))) %>% 
  expand_grid(age10 = unique(sealdat$age10)) %>% 
  mutate(year_fct = factor(year_fct, levels = levels(sealdat$year_fct))) %>% 
  mutate(predicted = predict(breeding_mod, 
                             newdata = ., 
                             re.form = ~ (1 | year_fct), type = "response"),
         age = age10 * 10)

breed_mod_summary <- summary(breeding_mod)

breed_mod_summary

#change rightmost value to say p value 
#add an extra sig fig to all columns
#p values are also rounded....

#table s1, look at roxannes ecology letters paper supplemental 
#Pull out the fixed effects table with 2 sfs, can paste in excel to adjust digits 
#intercept, age (linear), age (quadratic) as row names 
#in figure caption say "results from x analyis. The random effect variance of animal id is X (from results) and year was X. Apply to all the other stats. 
#analyses fit to the age in which we have larger sample sizes fit also have a quadratic fit and don't change the story

@@

Trying segmented analysis. Note: `segmented` does NOT support mixed effect GLMs!

```{r}
#| label: age_breeding_seg

breeding_mod2 <- glm(
  observed_int ~ age,
  data = sealdat, 
  family = "binomial"
)
breed_nseg <- lapply(
  list(15, c(9, 15)),
  \(.psi) {
    segmented(breeding_mod2, seg.Z = ~age, psi = .psi)
  }
)
AIC(breeding_mod2, breed_nseg[[1]], breed_nseg[[2]]) %>% 
  as_tibble() %>% 
  mutate(model = paste(1:3, "segments")) %>% 
  transmute(model, AIC, dAIC = AIC - min(AIC)) %>% 
  arrange(dAIC)

breeding_best <- breed_nseg[[1]]
breeding_pred <- tibble(age = seq(4, 22, length.out = 100)) %>% 
  predict(breeding_best, 
          newdata = ., 
          type = "response", 
          interval = "confidence") %>% 
  as_tibble() %>% 
  set_names(c("bred", "bred_lwr", "bred_upr")) %>% 
  mutate(age = seq(4, 22, length.out = 100))
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22)

ggplot(breeding_pred, aes(age)) +
  geom_ribbon(aes(ymin = bred_lwr, ymax = bred_upr),
              fill = "grey90") + 
  geom_line(aes(y = bred)) +
  geom_point(aes(y = perc_breed), breed_nonbreed) + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                breed_nonbreed,
                width = 0) +
  theme_classic()

@@

```{r}
#| label: fig-age_dist
#| fig-cap: "A: Age structure of northern elephant seals at Año Nuevo. B: Percentage of each age class that were observed breeding."
#| fig-width: 9

#Age distribution panel A

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

#Panel B: percentage of each age class that were breeders or  nonbreeders

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = year_fct),
            ranef_year,
            inherit.aes = FALSE,
            alpha = 0.2) +
  geom_text(aes(x = 22.1, predicted, label = year_fct),
            filter(ranef_year, age == 22),
            inherit.aes = FALSE,
            hjust = 0,
            alpha = 0.5,
            size = 3) +
  geom_line(aes(age, predicted),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) + 
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

plot_grid(age_dist, breed_nonbreed, nrow = 1, labels = "AUTO")

ggsave("outputs/agefig1.png", plot = , height = 4, width = 9)

Population level prediction - the fact that the raw data falls above it comes from the fact that those age classes are correlated with years and individuals that were very successful.

>>>>

First, address detection probability. DO THIS IN MARK.

@@

~~We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).~~

```{r}
#| label: breeding-mod

# Fit separate GLMMs to the data subsetted before and after onset of senescence
# We a priori chose age 12 for that (age_senesce)
# Pre-senescence model
breed_age_mod1 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    # Here's the age filter
    data = filter(sealdat, age < age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)
# Post-senescence model
breed_age_mod2 <- glmer(
    observed_int ~ age10 + (1 | animalID) + (1 | year_fct),
    data = filter(sealdat, age >= age_senesce),
    family = "binomial",
    control = glmerControl(optimizer="bobyqa")
)

tibble(
  model = c("Pre-senescence", "Post-senescence"),
  age_coef = c(fixef(breed_age_mod1)["age10"],
               fixef(breed_age_mod2)["age10"]),
  age_coef_se = c(summary(breed_age_mod1)$coef["age10", "Std. Error"],
                  summary(breed_age_mod2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(breed_age_mod1)$coef["age10", "z value"]),
                  pnorm(summary(breed_age_mod2)$coef["age10", "z value"]))
) %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3)))

@@

```{r}
#| label: fig-breeding-mod 
#| fig-width: 9
#| fig-height: 6

# Age distribution panel A

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

# Create POPULATION-LEVEL (i.e. without rand effects) predictions for both
# models and row bind them together
breed_pred <- rbind(
  # Pre-senescence
  ggpredict(
    breed_age_mod1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           # Add an age_class variable to facilitate plotting
           age_class = "young"),
  # Post-senescence
  ggpredict(
    breed_age_mod2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
) %>% 
  filter(age <= 20)

# Now create rand effect predictions
# This utility function pulls rand effect *coefs* from a model
get_ranefs <- function(model, r, reduce = TRUE) {
  result <- ranef(model)[[r]] %>% 
    as_tibble(rownames = r) %>% 
    rename(ranef = 2) %>% 
    arrange(ranef)

  if (reduce)
    slice(result, round(seq(1, nrow(result), length.out = 5)))
  else 
    result
}

# Get year and individual random effeccts from pre- and post-senescence models
year_ranef1 <- get_ranefs(breed_age_mod1, "year_fct")
year_ranef2 <- get_ranefs(breed_age_mod2, "year_fct")
indiv_ranef1 <- get_ranefs(breed_age_mod1, "animalID")
indiv_ranef2 <- get_ranefs(breed_age_mod2, "animalID")

# This util fun pulls rand effect *predictions* from a model
get_ranef_pred <- function(model, r, keep) {
  ggpredict(model,
            terms = c("age10 [all]", r), 
                             type = "random",
                             interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         "{r}" := group) %>% 
  semi_join(keep, by = r) %>% 
  filter(age <= 20)
}

# Predictions based on 5 different *years* (min, q1, median, q3, max ranef
# coef)
breed_pred_year <- rbind(
  # Pre-senescence
  get_ranef_pred(breed_age_mod1, "year_fct", year_ranef1) %>% 
    mutate(age_class = "young"),
  # Post-senescence
  get_ranef_pred(breed_age_mod2, "year_fct", year_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

# As above, but 5 different *individuals*
breed_pred_indiv <- rbind(
  get_ranef_pred(breed_age_mod1, "animalID", indiv_ranef1) %>% 
    mutate(age_class = "young"),
  get_ranef_pred(breed_age_mod2, "animalID", indiv_ranef2) %>% 
    mutate(age_class = "old")
) %>% 
  filter(age <= 20)

# The actual raw data
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = n(), 
            perc_breed = sum(observed_int) / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

# Figure with raw data and model predictions
breed_fig <- ggplot(breed_pred, aes(x = age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = interaction(year_fct, age_class)),
            breed_pred_year,
            linewidth = 0.75, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 3) +
  geom_text(aes(y = predicted, label = year_fct),
            filter(breed_pred_year, age == 4),
            x = 3.8,
            hjust = 1,
            size = 3) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 20.5)) +
  scale_y_continuous("Breeding %", 
                     limits = c(NA, 1), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))

age_dist | breed_fig

@@
